Wrattler: \textnormal{A platform for AI-assisted data science}

### 2018-02-20 report/paper.tex
Extend, copy paste AI assistant text, add Data Store thoughts

\section{Introduction}
Data science is an iterative, exploratory process that requires a collaboration between a 
computer system and a human. A computer can provide advice 
based on statistical analysis of the data and discover hidden structures or corner cases, but only
a human can decide what those mean.
A data science environment of the future thus needs to support an interaction model that 
keeps the human in the loop, allows an efficient human-computer interaction 
and can be extended with new AI asistants that provide advice about data.
In this report, we present Wrattler, a notebook system that is:

@@

\paragraph{Reproducible.} 
Wrattler separates the task of running scripts from the task of managing state.
A data store tracks the provenance and semantics of data, supports versioning and keeps 
the history, making the data analyses fully reproducible.

@@

\section{Wrattler and notebooks}

@@

@@

\begin{figure}
\vspace{-1em}
\includegraphics[scale=0.6]{diagram.pdf}
\vspace{-1em}
\caption{\small{In notebook systems such as Jupyter, state and execution is managed by a kernel. In
  Wrattler, those functions are separated and enriched with AI assistants.}}
\label{fig:arch}
\vspace{-1em}
\end{figure}

\section{Wrattler components}

@@

\begin{figure}
\includegraphics[scale=1,trim=0.5cm 0.5cm 0.5cm 0.5cm]{graph.pdf}
\vspace{-0.5em}
\caption{\small{Sample dependency graph. Data acquisiton and analysis are represented as 
opaque R/Python code cells. Data cleaning is written in a domain-specific langauge understood by
Wrattler and represented by multiple cells. Latter two cells depend on data frames exported by
earlier cells.}}
\label{fig:deps}
\vspace{-0.5em}
\end{figure}

\subsection{AI assistants}

@@

The above script is constructed interactively. When the user types the first line and types
``.'' the datadiff assistant offers the most likely patches and the user can choose. While
choosing, a preview of the result is computed on-the-fly, giving the user an immediate feedback.
Finally, the script is represented in a fine-grained way (second cell in Figure~\ref{fig:deps}).
Once it is interactively constructed, it can be translated to multiple supported languages
such as R, Python or JavaScript.
\end{document}

>>>>

\section{Introduction}
Data science is an iterative, exploratory process that requires a collaboration between a 
computer system and a human. A computer can provide advice based on statistical analysis of the 
data and discover hidden structures or corner cases, but only a human can decide what those mean
and decide how to handle them in data science scripts.

A data science environment of the future needs to support an interaction model that keeps the 
human in the loop, allows an efficient human-computer interaction and can be extended with new 
AI asistants that provide advice about data. Such advice needs to be explainable to the human user,
as well as machine-readable by other components of the system. In this report, we present Wrattler, 
a system that is:

@@

\paragraph{Reproducible.} 
Wrattler separates the task of running scripts from the task of managing state. 
This is handled by a data store, which tracks the provenance and semantics of data, supports 
versioning and keeps the history, making the data analyses fully reproducible.

@@

\paragraph{Explainable.}
The hints provided by AI assistants are explainable. Rather than working as black boxes that 
transform one dataset into another, AI assistants generate scripts in simple domain-specific
languages, that specify how the data should be transformed. Those scripts can be reviewed 
and modified by a human.

\vspace{0.5em}
\noindent
In the rest of this document, we discuss limitations of current notebook systems and how Wrattler
resolves them (Section~\ref{sec:overview}). We discuss how the individual components of Wrattler
work together (Section~\ref{sec:wrattler}) and then focus on two of them in detail --
we look at how Wrattler can be extended with AI assistants (Section~\ref{sec:ai}) and
how semantic information about data is managed by the data store (Section~\ref{sec:datastore}).

\section{Wrattler and notebooks}
\label{sec:overview}

@@

\begin{figure}
\vspace{-0.5em}
\includegraphics[scale=0.6]{diagram.pdf}
\vspace{-0.5em}
\caption{\small{In notebook systems such as Jupyter, state and execution is managed by a kernel. In
  Wrattler, those functions are separated and enriched with AI assistants.}}
\label{fig:arch}
\vspace{-0.5em}
\end{figure}

@@

\section{Wrattler components}
\label{sec:wrattler}

@@

\begin{figure}
\includegraphics[scale=1,trim=0.5cm 0.5cm 0.5cm 0.5cm]{graph.pdf}

\caption{\small{Sample dependency graph. Data acquisiton and analysis are represented as 
opaque R/Python code cells. Data cleaning is written in a domain-specific langauge understood by
Wrattler and represented by multiple cells. Latter two cells depend on data frames exported by
earlier cells.}}
\label{fig:deps}
\vspace{-0.5em}
\end{figure}

\subsection{AI assistants}
\label{sec:wrattler-ai}

@@

The above script is constructed interactively. When the user types the first line and types
``.'' the datadiff assistant offers the most likely patches and the user can choose. While
choosing, a preview of the result is computed on-the-fly, giving the user an immediate feedback.
Finally, the script is represented in a fine-grained way (second cell in Figure~\ref{fig:deps}).
Once it is interactively constructed, it can be translated to multiple supported languages
such as R, Python or JavaScript.

\section{AI assistants}
\label{sec:ai}

{\color{blue}
An AI assistant guides data analysts through a single data collection, integration, preparation, 
analytics, or reporting task. Typically analysts will interact with many separate assistants over 
the course of an analysis, each of which specialize in a single kind of task.
}
\subsection{Architecture of AI assistants}

{\color{blue}
AI assistants have access to information in the Wrattler data store and can use two separate
aspects for their guidance:
%
\begin{itemize}
\item[--] AI assistants can access all data stored in the data
store. This includes imported external data files, such as raw text or raw HTML content from which
the analyst wants to extract data, as well as structured data frames imported from CSV files or
produced as a result of previous computations. Moreover, an AI assistant can also access multiple
versions of a file.
\vspace{0.25em}

\item[--] AI assistants can access the dependency graph created for a 
notebook. This allows the them to access the transformation history, which is the code 
that has already been run during the analysis. The idea is that assistants may need to know the 
context of transformations that have already been performed in order to decide what to suggest next.
\end{itemize}

The assistants should be self-contained in the sense that they interact only through the 
data store. This restriction is designed to encourage the Unix "do one thing well" design for 
assistants.

There are two important design aspects of AI assistants. First, they always produce code, rather
than being a black-box that transforms data. Second, they are designed to be used interactively.

\paragraph{Domain specific languages.}
The output of an assistant is code in a small domain-specific language designed for the task that
the AI assistant is solving. The code can be used in two ways -- during the interaction with the
AI assistant, the code is exposed through a Wrattler DSL framework as discussed in 
Section~\ref{sec:wrattler-ai}. For production use, the code is translated to a language like R or 
Python. A typical code produced by an AI assistant will implement a data transformation or
produce a visualization.

\paragraph{Interactivity.}
Assistants should usually contain user interaction in their design, and adapt rapidly to feedback 
from the analyst. For example, a data cleaning assistant would probably want to ask the user to 
confirm potential changes, and if the user says no, to reconfigure the cleaning module 
significantly so that the user does not need to keep rejecting similar changes.

However, at the same time, our data preparation tools should be usable in "batch mode" by analysts 
who already know what they want to do and don't want to mess with a wizard. We should aim to 
produce the equivalent of scikit-learn for data preparation.

The output snippet should be readable, and something that a data analyst can look at to verify 
that the transformation accomplishes what they had wanted. As far as possible, the snippet 
should be "reusable" in a sense that it could be copy-pasted into another script that was 
processing a similar dataset and would still work. You can think of a snippet as something 
that a person might write in a single cell of a Jupyter notebook.
}

\subsection{Types of AI assistants}
{\color{blue}
AI assistants can be imagined for every step of the data analytics process. We attempt a fairly c
omprehensive list of the data analytics tasks for which can imagine targetting assistants. Then, 
for each task, we list a few examples of specific agents. Even though the list of agents is 
incomplete, it is still too long to implement fully in the current project.

\paragraph{Data collection.} Assistants can search the data store, as well as external repositories
to recommend data sets that are relevant to the task at hand and offer to import these. For example,
when analysing data about different countries, more information about countries can be offered.

\paragraph{Data integration.} Assistants can identify pairs of variables across multiple data 
sources that are likely to be coding the same variable.

\paragraph{Schema inference.} Inferring types of variables and doing this probabilistically.

\paragraph{Data parsing.}
Quick tool for specifying a scraper for web pages (related to wrapper induction, Sumit Gulwani 
work, possibly this paper Robust and Noise Resistant Wrapper Induction
Assistant to automatically infer format of csv files
Assistants to extract tables from pdf's, html
Identify which values are likely to be "special", like missing data markers

\paragraph{Data cleaning}
Identifying values that seem to be out of range, e.g. missing value indicators
Identifying possible inconsistent coding of strings
Identifying potential functional dependencies (1:1 correspondences) and returning violations
Manual cleaning of subset of data (a la Sample Clean)
Indentifying rows that seem "jointly" unusual, i.e., that seem not to conform to statistical dependencies among the columns
Identifying distinct records that may refer to the same entity (record linkage)

\paragraph{Data monitoring.}
Data diff: Returning a report of the differences between two data sets
Monitoring classifier to detect shifts in its input distribution
Detecting change points and covariate shift
Montioring classifier to suggest when/how often new data points shoud be labelled for monitoring

\paragraph{Exploratory visualization.}
Summarizing most representative rows of a data table
Automatically styling scatterplots and bar charts

\paragraph{Performing analytics.}
Which classification methods should I use? e.g., by search through similar problems on Kaggle and Open ML (a la AutoML)
Simiarly for feature selection methods and so on
Hyperparameter exploration

\paragraph{Debugging and interpreting predictive analytics.}
Debugging predictive analytics, e.g., deciding whether and how to add more features
Error analysis assistant: Summarize the output on validation data
Error analysis based on visualization of classifier score (see work of Saleema Amershi)
Tools for explaining the predictions of a model on validation data (a la LIME) * Research directions: Improved model families for the interpretable explanation. Decision lists on marginal distribution rather than joint
"Training set blame": Given an error in the validation data, what examples from the training set caused me to make that error?
Explaining differences between predictions: why was instance 24601 treated differently than 24675?
}
\subsection{Possible AI assistants}

{\color{blue}

\paragraph{Data Parsing.}

Simple scraper assistant: Simple information extraction wizard. Create a data table based
on HTML pages that have been automatically generated, like product review pages,
for which a simple pattern on the HTML tree will be sufficient to extract the desired information.
This is called wrapper induction in the information extraction literature.

* *Input*: Web page that contains information in unstructured form, such as a list of search resuts or an HTML list.

* *Output*: A data frame (i.e. a relational table) that contains the information that has
been extracted from the web page.

* *User interaction*: User needs to initially highlight and specify which items to be scraped.
Need to summarize the results of the scraping and highlight cases
which might require correction from the user. Update scraper accordingly.

* *Potential method*: A wrapper is a path through the HTML DOM. Use a language (like XPath) for navigating through the DOM. Search for the shortest path that retrieves
the labels provided by the user.

* *Research issues*: Not clear, given the large amount of related work in this area.
We would need to read around carefully before we can identify opportunities.
Possible opportunities include:
    (a) Using a language like [Scrapy](https://scrapy.org/) to induce the wrappers;
    (b) Using a language model over the scraped values to fine-tune the wrapper (e.g. if a wrapper returns both dates and cities, it's probably wrong), or
	(c) using a probabilistic model over the program representation of the wrapper to focus attention in the search
for good wrappers. Perhaps there are enough scrapers on line (e.g. Github projects that use scrapy) that you can get a training set for this.
* *Related work*: Gulwani's work on data wrangling, lots of work on wrapper induction.
There is a lot of work on this. Scrapy is a popular open source framework that several
of my BSc/MSc students have found useful.

\paragraph{Data cleaning.}
*Inconsistent string assistant*: Identify potential typos in string data. The idea is to look for strings that do not occur commonly in the data set, but are very close in edit distance
to strings that do occur commonly.

* *Input*: A column of a data table whose values are strings, like addresses, cities, states/provinces, occupations, and so on.

* *Output*: A list of suggestions for strings that should be renamed: e.g. `[ ("Stocland", "Scotland"), ("Untied Kringdom", "United Kingdom") ]`

* *Potential method*: Simplest cut: Define a threshold $\delta$ on Levenshtein distance,
and $0 < \alpha < 1$ on the number of occurrences. Return the set of all strings $y$ that
occur less than $\alpha N$ times in the data set but where there is a corrected string $x$ which is less than $\delta$ in string edit distance from $y$.
More ambitious: Noisy channel model for correction.
Learn a probability distribution $p(x)$ over the strings in the column,
and  define an error model $p(x'|x)$. Then search for ways to transform the  dirty column $x'$  to a clean version $x$ in such a way to maximize $p(x|x')$.
This formulation deserves some more thought,
though, before we try to implement it.

* *Research questions*: Need to connect this to existing concepts in the databases and
data mining literature, such as U-repair.

* *User Interaction*: The user should be able to reject suggestions in the output list,
and this should rapidly update the models used in such a way that similar suggestions
to the bad ones are also removed from the list. A noisy channel model would be nice
for this, not sure about the edit distance one.

\paragraph{Exploratory visualization.}

*Data table summarization assistant*: Displays a short subtable of the most "representative"
columns of a given data frame. The number of rows in the summary should be small enough
to fit on a given screen. Essentially an alternative to the `head()` function in pandas/R.

*Input*:  A data frame, along with type information (continuous/categorical) for each of the columns. Desired number $k$ of rows in output.

*Output*: A list of $k$ indices of rows in the dataframe

*Potential method*:  Simplest possible implementation:
Run $k$-medoids and return the medoid values.

*Research questions:* First, are there research questions here, or is this so simple
that it  is part of the infrastructure?  If $k$-medoids is so simple for this,
why doesn't everyone also do it? How to evaluate whether the summary is good?
Maybe there are examples of data sets where medoids is not good.
For example, in the Karpathy ICML example, medoids would not make sense,
listing the institutions that have the most papers is most informative, because
it's a "long tail" type of column. How do you tell "long tail" data from "just show
the clusters" type data? Perhaps a model-based framework could distinguish?
Maybe you want to summarize the ways in which two data sets differ?
Is the best summary the cluster centroids or the "top k" along some value? 
* *User interaction:* I'm not sure what interactions are enecessary.
Users could ask for more rows, mark two items as similar, or mark an item as uninteresting.
Or perhaps better would be to allow drill down, i.e., to make it easy to explore
the clusters represented by each example.
}

\section{Data store}
\label{sec:datastore}

{\color{red}
\textbf{TP: My rough notes from discussion with JG}

~

\noindent
Data store stores data frames with metadata. This is a good start for data science work,
because most languages can provide mappings to/from data frames.
}

\subsection{Semantics annotations}
{\color{red}
Data store provides a way of annotating cells, columns, combinations of columns and rows.
For example, a column may be annotated with a type that indicates that its values are a mix of
postcodes and city names. A cell can be annotated with an information saying that this value is 
more likely a postcode than a city name. We also need to support annotaitons on a combination of
columns, because three columns can jointly represent an address. A row can be annotated with a 
provenance -- for example, when merging rows from two data sets with different sources.
}

\subsection{Annotation format}
{\color{red}
We want a lightweight annotation format so that implementing an AI assistant or a language
runtime that needs to communicate with the data store is not too hard.
Our annotation format will follow something like \url{http://schema.org}, which defines
a hierarchy of entity types and has a simple way for adding annotations to things.
We will need to define our own hierarchy for schemas for data science.
}

\subsection{Schemas for data science}
{\color{red}
This hierarchy will include some basic things that the data store understands and can 
convert between (integers, floats, unbounded size integers, ISO dates, etc.)

In addition, we will define some purely semantic annotations that are commonly useful and 
we will define a hierarchy for those. This represents things like addresses, postcodes, countries,
cities, etc.

Everyone can define their own schemas (like schema.org) by defining their own namespace (URL)
and providing entities with their custom namespace. The data store will simply save and return
this information - without doing any checking.}

\subsection{Content negotiation}
{\color{red}
The only bit where the data store does something clever is that it will be able to return data
in a format that the client asked for (a bit like HTTP negotiation). The client can say, ``I only
understand semantic information defined by these schema namespaces`` and the data store will do
its best to return data in the required format.

For simple types (e.g. dates) that the data store understands, it will convert data accordingly.
For other semantic information that the data store does not understand (e.g. information that
a certain cell is missing with a probability $p$), the data store will filter out the extra
information and just return the raw number.
}

\end{document}
